---
title: "Breast Cancer Classification"
author: 
  - Parviz Hajizada
  - Duy Tuan Doan
output: html_document
---

# Introduction
Breast Cancer is one of most earliest cancer recorded in human history and have taken countless lives. Even though no reliable and permanent cure have not been developed, detection at early stages is crucial because the sooner the better.

In this paper, we build classification models to identify if the tumor is benign (harmless) or malignant (harmful). This dataset is collected and provided by University of Wisconsin (1995) [link](ftp://ftp.cs.wisc.edu/math-prog/cpo-dataset/machine-learn/cancer/). It is composed of 30 descriptive features that are from a digitized image of a fine needle aspirate (FNA) of a breast mass.

#### Load necessary packages
```{r message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(tibble)
library(corrplot)
library(ggplot2)
library(AER)
library(lmtest)
library(nnet)
library(MASS)
library(caret)
library(verification)
library(e1071)
library(janitor)
library(class)
```

#### Import data
```{r message = FALSE}
wdbc <- read_csv("wdbc.csv")
```

## Data Preprocessing
#### Rename the columns
```{r}
names(wdbc) <- c("radius1", "texture1", "perimeter1", "area1", "smoothness1", "compactness1", "concavity1", "concave_points1", "symmetry1", "fractal_dimension1",
                 "radius2", "texture2", "perimeter2", "area2", "smoothness2", "compactness2", "concavity2", "concave_points2", "symmetry2", "fractal_dimension2",
                 "radius3", "texture3", "perimeter3", "area3", "smoothness3", "compactness3", "concavity3", "concave_points3", "symmetry3", "fractal_dimension3",
                 "class")
```

#### Factorize target variable
```{r}
# b stand for benign, m stands for malignant type of tumor 
wdbc$class <- factor(wdbc$class, levels = c(1, 2), labels = c("b", "m"))
```

## Exploratory data analysis
#### Missing data
```{r}
any(is.na(wdbc)) 
```

#### Summary statistics
```{r}
summary(wdbc) 
```

#### Distribution of each variable
```{r pressure, echo=FALSE,out.width = "110%", out.height='150%'}
par(mfrow=c(3,3))
#convert as data frame 
wdbc <- as.data.frame(wdbc) 
#simple plot for each column for quick checking 
 
for (i in seq(1, length(wdbc),1)){ 
  if (class(wdbc[,i]) == 'numeric'){ 
    hist(wdbc[,i], main= names(wdbc)[i]) 
  } else if (class(wdbc[,i]) == 'factor'){ 
    barplot(table(wdbc[,i]), main= names(wdbc)[i]) 
  } else {  
    print('there is nothing to plot') 
    } 
} 
par(mfrow=c(1,1)) 
```

Majority of the distributions are right skewed and the target variable is classified as benign in most of the cases. So it is better not to use accuracy but ROC to quantify the performance of the models. 

#### Structure of the data
```{r}
sapply(wdbc, class)
```

# Modelling
#### Data Partitioning
```{r}
set.seed(1)
which_train <- createDataPartition(wdbc$class, 
                                   p = 0.8, 
                                   list = FALSE) 
```

#### Train and test set
```{r}
train <- wdbc[which_train,]
test <- wdbc[-which_train,]
```

#### Distribution of the dependent variable in both sets
```{r}
tabyl(train$class)
tabyl(test$class)
```

#### Cross validation
```{r}
ctrl_cv5x3 <- trainControl(method = "repeatedcv",
                           number = 5,
                           classProbs = TRUE,
                           summaryFunction = twoClassSummary,
                           repeats = 3)
```

## Logistic Regression
```{r, warning=FALSE}
logit.train <- 
  train(class ~ ., 
        data = train, 
        method = "glm",
        metric = "ROC",
        family = "binomial",
        trControl = ctrl_cv5x3)
```

#### Fitted values
```{r}
logit.train_fitted <- predict(logit.train,
                              train,
                              type = "prob")
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = as.factor(ifelse(logit.train_fitted["b"] > 0.5, "b", "m")), 
                reference = train$class, 
                positive = "b") 
```

#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         logit.train_fitted[,"b"])$A
```

#### Predicted values
```{r}
logit.train_forecasts <- predict(logit.train,
                                 test,
                                 type = "prob")
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = as.factor(ifelse(logit.train_forecasts["b"] > 0.5, "b", "m")), 
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.logit <- roc.area(ifelse(test$class == "b", 1, 0),
                      logit.train_forecasts[,"b"])$A
```

## Linear Discriminant Analysis
```{r}
lda.train <- 
  train(class ~ ., 
        data = train, 
        method = "lda",
        metric = "ROC",
        trControl = ctrl_cv5x3)
```

#### Fitted values
```{r}
lda.train_fitted <- predict(lda.train,
                            train,
                            type = "prob")
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = as.factor(ifelse(lda.train_fitted["b"] > 0.5, "b", "m")), 
                reference = train$class, 
                positive = "b") 
```

#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         lda.train_fitted[,"b"])$A
```

#### Predicted values
```{r}
lda.train_forecasts <- predict(lda.train,
                               test,
                               type = "prob")
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = as.factor(ifelse(lda.train_forecasts["b"] > 0.5, "b", "m")), 
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.lda <- roc.area(ifelse(test$class == "b", 1, 0),
                    lda.train_forecasts[,"b"])$A
```

## Quadratic Discriminant Analysis
```{r}
qda.train <- 
  train(class ~ ., 
        data = train, 
        method = "qda",
        metric = "ROC",
        trControl = ctrl_cv5x3)
```

#### Fitted values
```{r}
qda.train_fitted <- predict(qda.train,
                            train,
                            type = "prob")
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = as.factor(ifelse(qda.train_fitted["b"] > 0.5, "b", "m")), 
                reference = train$class, 
                positive = "b") 
```
 
#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         qda.train_fitted[,"b"])$A
```

#### Predicted values
```{r}

qda.train_forecasts <- predict(qda.train,
                               test,
                               type = "prob")
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = as.factor(ifelse(qda.train_forecasts["b"] > 0.5, "b", "m")), 
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.qda <- roc.area(ifelse(test$class == "b", 1, 0),
                    qda.train_forecasts[,"b"])$A
```

## KNN
```{r}
train.knn_cv_scaled <- 
  train(class ~ .,
        data = train, 
        method = "knn",
        trControl = ctrl_cv5x3,
        tuneGrid = data.frame(k = 1:30),
        metric = "ROC",
        preProcess = c("range"))

```

#### Elbow curve
```{r}
plot(train.knn_cv_scaled) 
```

#### Fitted values
```{r}
knn.fitted <- predict(train.knn_cv_scaled,
                      train)
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = knn.fitted,
                reference = train$class,
                positive = "b")
```

#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         ifelse(knn.fitted == "b", 1, 0))$A
```

#### Predicted values
```{r}
knn.forecasts <- predict(train.knn_cv_scaled,
                         test)
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = knn.forecasts,
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.knn <- roc.area(ifelse(test$class == "b", 1, 0),
                    ifelse(knn.forecasts == "b", 1, 0))$A
```

## SVMLinear
#### Parameters
```{r}
modelLookup("svmLinear") 
```

#### Grid of parameters
```{r}
parametersC <- data.frame(C = c(0.001, 0.01, 0.02, 0.05,  
                                0.1, 0.2, 0.5, 1, 2, 5))
```

#### Train data
```{r}
svm_Linear <- train(class ~ .,  
                    data = train,  
                    method = "svmLinear", 
                    tuneGrid = parametersC, 
                    metric = "ROC",
                    trControl = ctrl_cv5x3) 
```

#### Predicted values
```{r}
svm_Linear_forecasts <- predict(svm_Linear,  
                                newdata = test) 
```

#### Fitted values
```{r}
svm_linear_fitted <- predict(svm_Linear,
                             train)
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = svm_linear_fitted,
                reference = train$class,
                positive = "b")
```

#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         ifelse(svm_linear_fitted == "b", 1, 0))$A
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = svm_Linear_forecasts,
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.svm_linear <- roc.area(ifelse(test$class == "b", 1, 0),
                           ifelse(svm_Linear_forecasts == "b", 1, 0))$A
```

## SVMRadial
#### Parameters
```{r}
modelLookup("svmRadial") 
```

#### Grid of parameters
```{r}
parametersC_sigma <-  
  expand.grid(C = c(0.01, 0.05, 0.1, 0.5, 1, 5), 
              sigma = c(0.05, 0.1, 0.2, 0.5, 1)) 

```

#### Train data
```{r warning = FALSE, message = FALSE, include = FALSE}
svm_Radial <- train(class ~ .,  
                    data = train,  
                    method = "svmRadial", 
                    tuneGrid = parametersC_sigma, 
                    trControl = ctrl_cv5x3) 
```

#### Predicted values
```{r}
svm_Radial_forecasts <- predict(svm_Radial,  
                                newdata = test) 
```

#### Fitted values
```{r}
svm_Radial_fitted <- predict(svm_Radial,
                             train)
```

#### Confusion matrix - train set
```{r}
confusionMatrix(data = svm_Radial_fitted,
                reference = train$class,
                positive = "b")
```

#### ROC - train set
```{r}
roc.area(ifelse(train$class == "b", 1, 0),
         ifelse(svm_Radial_fitted == "b", 1, 0))$A
```

#### Confusion matrix - test set
```{r}
confusionMatrix(data = svm_Radial_forecasts,
                reference = test$class, 
                positive = "b") 
```

#### ROC - test set
```{r}
roc.svm_radial <- roc.area(ifelse(test$class == "b", 1, 0),
                           ifelse(svm_Radial_forecasts == "b", 1, 0))$A
```

## Choose the best model
#### ROC
```{r}
test.roc <- data.frame(model=c("logit","lda","qda","knn","svm_linear", "svm_radial"),
                       roc=c(roc.logit, roc.lda, roc.qda, roc.knn,roc.svm_linear, roc.svm_radial))
test.roc <- test.roc[order(test.roc$roc, decreasing=TRUE),]
test.roc$model <- factor(test.roc$model, levels=test.roc$model)
test.roc
```

#### Plot AUC
```{r}
theme_set(theme_gray(base_size = 18))
ggplot(data = test.roc, aes(x = model, y = roc)) +
  geom_bar(stat="identity")
```